openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2022-03-22 paper/main.tex
Update benchmarking figures in manuscript

However, easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many theoretical advancements, the most popular t-SNE libraries have been slow to incorporate them into their implementations. In this paper, we present \opentsne, our open-source Python implementation of t-SNE, and the recently proposed extensions to t-SNE that \opentsne\ implements. \opentsne\ is easy to install and includes precompiled binaries available through popular Python package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

\citet{belkina2019automated} later found that convergence can be sped up by
increasing the learning rate from the standard $\eta=200$ to $\eta=N/12$. As a
side-effect, embeddings converge faster, and the number of iterations can be
lowered from the typical 1000 to 750, decreasing the overall runtime.  Modern
t-SNE implementations, including \pkg{FIt-SNE} and \opentsne, have adopted this
convention.

@@

\subsubsection*{Repulsive Forces}
Similarly, the repulsive term can also be approximated, motivated by methods from
particle simulations. \citet{van2014accelerating} proposed an approach based on
N-body simulations and used a space-partitioning Barnes-Hut tree approach to
approximate the interaction between data points. This reduces the time
complexity from $\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$. More recently,
\citet{linderman2019fast} proposed an alternative approach, FIt-SNE, based on
non-uniform convolutions which further reduces the time complexity to
$\mathcal{O}(N)$.

@@

t-SNE is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Therefore embeddings of new data
points need to be found through the use of optimization
techniques~\citep{policar2021embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place
while new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \qquad
q_{j \mid i} = \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}, \notag
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eqn.~(\ref{eq:symmetrize}). Plugging these terms into
Eqn.~(\ref{eq:kl_divergence}), we obtain the following gradient
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}. \notag
\label{eq:gradient}
\end{equation}

@@

One of t-SNE's criticisms is that it fails to preserve long-range distances but instead focuses on capturing the local structure of the data manifold~\citep{becht2019dimensionality}. Recently, several approaches have been proposed to improve the global organization of the resulting embeddings.

@@

In standard t-SNE, distances between data points are converted to similarities through the use of Gaussian kernels. The perplexity of these kernels governs the sizes of the local neighborhoods which are to be preserved. One easy way to improve the global consistency of resulting embeddings is to increase the sizes of these neighborhoods. This can be achieved by increasing the perplexity parameter. However, this often comes at the price of local structure. \citet{kobak2019art} propose to use mixtures of Gaussians instead to preserve better short-range and long-range distances and, thus, achieve a better trade-off between local and global structure.

Embeddings produced by standard t-SNE often use all available space and separate clusters by only thin boundaries. When working with large data sets, this often obscures the global relationships between clusters as all neighboring groups appear equidistant from one another.  Other dimensionality reduction methods, such as UMAP and ForceAtlas2, produce embeddings where clusters appear more compact, and the white space separating the clusters may be interpreted as a loose measure of distance. Recently, \citet{bohm2020unifying} showed that the exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP and ForceAtlas2. By incorporating exaggeration into later phases of the optimization, t-SNE introduces white space between clusters, which better reflects the global relations between clusters.

@@

\subsection{Structure of openTSNE}

\opentsne\ is designed to accomodate all users, including both a simple API that allows users to quickly obtain standard t-SNE visualization, as well as an advanced API that allows practitioners and researchers to modify different aspects of the t-SNE algorithm.

The t-SNE algorithm is dependent on three key steps: the initial positions of points in the embedding, an affinity model describing the similarities between data points in the ambient space, and the optimization which optimizes the embedding layout. \opentsne\ mirrors these different components in its design. \opentsne\ contains three corresponding core modules -- \code{initialization}, \code{affinity}, and \code{tsne} -- each of which will be decribed in detail in later sections. These modules are designed to be easily extensible, allowing researchers to experiment with different configurations and produce the best possible t-SNE visualizations. For less demanding users, \opentsne\ also provides a familiar, scikit-learn-like interface through the \code{openTSNE.TSNE} class.

\subsubsection{Initialization}

just a numpy array

\subsubsection{Affinities}

\paragraph{Affinity}

\paragraph{KNNIndex}

\subsubsection{Embedding objects}

\paragraph{TSNEEmbedding}

\paragraph{PartialTSNEEmbedding}

\subsection{Old text}

We introduce \opentsne, a comprehensive Python library that implements t-SNE and its numerous recently proposed extensions, including the implementation of efficient approximation schemes allowing the embedding of millions of data points~\citep{van2014accelerating,linderman2019fast}, the addition of new data points into existing embeddings~\citep{policar2021embedding}, various initialization improvements~\citep{kobak2019umap}, extensions~\citep{kobak2019heavy,kobak2019art}, and improved parameter defaults~\citep{belkina2019automated}. When combined, these improvements result in the faster generation of more informative visualizations.

\opentsne\ is compatible with the Python data science ecosystem (e.g. \pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar and intuitive API to users. Its modular design encourages extensibility and experimentation with various parameter settings and changes to the analysis pipeline. \opentsne\ aims to make the latest theoretical advances in t-SNE accessible to the wider data science community. These features are easily accessible through an intuitive API, which closely follows the style \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit for machine learning. The following snippet exemplifies this interface and demonstrates how \opentsne\ can be used to quickly generate several visualizations, each highlighting different aspects of the underlying data structure.
\begin{CodeChunk}
\begin{CodeInput}
from openTSNE import TSNE

@@

\section{Summary and discussion} \label{sec:summary}

@@

When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between clusters become increasingly difficult to interpret. We constructed Fig.~\ref{fig:cao}.a from the data containing expression profiles of over two million single cells captured at different time points in mouse development. The embedding indicates numerous clusters with transitions between time points, as marked by the color-coding, that are difficult to interpret. Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently, more informative visualizations~\citep{kobak2019art}. For instance, \ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the data's overall structure. Further doubling the exaggeration in \ref{fig:cao}.c allows us to observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters. The visualization also reveals several tiny clusters, possibly corresponding to rare cell types.
\begin{CodeChunk}
\begin{CodeInput}
# To improve the global structure of the final layout, we initialize the
# embedding by creating a t-SNE plot on a sample of points.

@@

# We can next use the `transform` functionality to determine starting points for
# the remaining data points. Since we don't actually want to optimize the
# remaining points separately from the sample points, we don't need to call the
# `transform` function. Instead, we'll call the `prepare_partial` function,
# which `transform` calls anyways under the hood.
init_rest = init_sample.prepare_partial(x_rest)

# Lastly, we merge `init_sample` and `init_rest`
init_full = np.vstack((init_sample, init_rest))[reverse]
\end{CodeInput}
\end{CodeChunk}
We generate the four different embeddings using the following code snippet (which is shortened for clarity).
\begin{CodeChunk}
\begin{CodeInput}
# We separately precompute the affinities as this can take quite a long time.
# This way, we can reuse the precomputed affinities in various different
# embeddings, without having to repeat the same, costly computation every time.
affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)

@@

% General problem statement
One of the t-SNE's common criticisms is limited scalability to large data sets
containing, for instance, millions of data
points~\citep{becht2019dimensionality}. The culprit for slow response time stems
from particular optimization procedures and their specific implementation in
popular open-source libraries. Until quite recently, most popular implementations of
t-SNE were based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$, or the Barnes-Hut approximation scheme developed by
\citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}(N \log
N)$, where $N$ is the number of data items ({\em e.g.} cells).

% Problem with Python
In Python, the most
widely-used implementation of t-SNE came from \pkg{scikit-learn}, which exhibits
long runtimes when compared to its C++ counterpart -- \pkg{MulticoreTSNE}. The
multi-threaded \pkg{MulticoreTSNE} implementation~\citep{Ulyanov2016} can
construct t-SNE embeddings of millions of data points in a matter of hours on
widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks_py}).
However, recently, \citet{linderman2019fast} developed a new approximation
scheme -- FIt-SNE -- which reduces the asymptotic time complexity to
$\mathcal{O}(N)$. This approximation scheme is included in \pkg{FIt-SNE} and \opentsne, enabling the embedding of large data sets in a matter of minutes.

% We benchmark Python
Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python t-SNE implementations,
including those from \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1),
\pkg{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.6.1). We perform benchmarks on
a server-grade Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original
t-SNE parameters, as some implementations do not allow their modification.
To generate benchmark data sets of different sizes, we subsampled data from the 10X Genomics 1.3 million mouse brain data set five times, resulting in five different data sets for each size. In total, we run each implementation on 30 different data sets.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{benchmarks_python}
  \caption{\label{fig:benchmarks_py}
  We benchmark \opentsne\ (v0.6.1) against three popular open-source
  implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
  (v1.0.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
  on a server-grade Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.Notice
  that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
  same interpolation-based approximation scheme, while \pkg{scikit-learn} and
  \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
}
\end{figure*}

@@

% We benchmark other programming languages
Using the same benchmarking setup as before, we ...
The results are shown in Fig.~\ref{fig:benchmarks_lang}.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{benchmarks_langs}
  \caption{\label{fig:benchmarks_lang}
  We choose the fastest available, open-source implementations of the t-SNE algorithm from the Python, R, and Julia programming languages. We choose \opentsne\ for Python, \pkg{Rtsne} for R, and \pkg{TSne.jl} for Julia. The benchmarks reflect the time complexity of the implemented algorithms. For instance, \pkg{TSne.jl} implements a naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all but the smallest data sets. \pkg{Rtsne} scales similarily to \pkg{MulticoreTSNE} in Python, as it implements the $\mathcal{O}(N \log N)$ Barnes-Hut approximation scheme. Interestingly, the multi-threaded implementation version of \pkg{Rtsne} crashes on benchmark data sets containing over 500,000 data points due to lack of memory. This would seriously hamper its utility on consumer-grade platforms, as our benchmark platform contained 128 GB of memory. \opentsne\ is the only library implementing the more recent FIt-SNE approximation scheme, which makes it suitable for even the largest data sets.
}
\end{figure*}

@@

We designed \opentsne\ to be accessible to a broader audience. We provide
precompiled binaries for all major Python versions on all major platforms,
making the installation process as seamless as possible. One can install
\opentsne\ through the Python Package Index (\pkg{PyPI}) or \pkg{conda} from the
\pkg{conda-forge} channel, the two most widely adopted Python package managers.
\opentsne's interface is inspired by \pkg{scikit-learn}, which is well
established in the Python data science ecosystem. \opentsne\ implements
multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation
schemes, enabling it to be applied to data sets containing millions of data
points. While the Python virtual machine inevitably introduces some performance
overhead, the runtime is comparable to its C++ counterpart -- \pkg{FIt-SNE}.
Finally, \opentsne\ is extensible. Its modular design enables researchers to
quickly experiment with different parameter settings and easily incorporate
custom components into the software. We provide an extended list of features and
comparison to other popular t-SNE implementations in Table~\ref{tab:features}.

>>>>

However, easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations. In this paper, we present \opentsne, our open-source Python implementation of t-SNE, complete with its recently proposed extensions. \opentsne\ is easy to install and includes precompiled binaries available through popular Python package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

\citet{belkina2019automated} later found that convergence can be sped up by
increasing the learning rate from the standard $\eta=200$ to $\eta=N/12$. As a
side-effect, embeddings converge faster, and the number of iterations can be
lowered from the typical 1000 to 750, decreasing the overall runtime.

@@

\subsubsection*{Repulsive Forces}
Similarly, the repulsive term can also be approximated, motivated by methods from
particle simulations. \citet{van2014accelerating} proposed an approach based on
N-body simulations and used a space-partitioning Barnes-Hut tree approach to
approximate the interaction between data points. This reduces the time
complexity from $\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$. More recently,
\citet{linderman2019fast} proposed an alternative approach, FIt-SNE, based on
non-uniform convolutions, which further reduces the time complexity to
$\mathcal{O}(N)$.

@@

The t-SNE algorithm is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Therefore embeddings of new data
points need to be found through optimization
techniques~\citep{policar2021embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place
while new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \qquad
q_{j \mid i} = \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}, \notag
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eqn.~(\ref{eq:symmetrize}). Plugging these terms into
Eqn.~(\ref{eq:kl_divergence}), we obtain the following gradient
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}. \notag
\label{eq:gradient}
\end{equation}

@@

One of t-SNE's long-standing criticisms is that it fails to preserve long-range distances but instead focuses on capturing the local structure of the data manifold~\citep{becht2019dimensionality}. Recently, several approaches have been proposed to improve the global organization of the resulting embeddings.

@@

In standard t-SNE, distances between data points are converted to similarities through the use of Gaussian kernels. The perplexity of these kernels governs the sizes of the local neighborhoods which are to be preserved. One easy way to improve the global consistency of resulting embeddings is to increase the sizes of these neighborhoods. This can be achieved by increasing the perplexity parameter. However, increasing perplexity often leads to loss in local sturcture and tends to obscure small clusters. \citet{kobak2019art} instead propose to use mixtures of Gaussians to preserve better short-range and long-range distances and, thus, achieve a better trade-off between local and global structure.

Embeddings produced by standard t-SNE often use all available space and separate clusters by only thin boundaries. When working with large data sets, this often obscures the global relationships between clusters as all neighboring groups appear equidistant from one another.  Other dimensionality reduction methods, such as UMAP and ForceAtlas2, produce embeddings where clusters appear more compact, and the white space separating the clusters may be interpreted as a loose measure of distance. Recently, \citet{bohm2020unifying} showed that the exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP and ForceAtlas2. By incorporating exaggeration into later phases of the optimization, t-SNE introduces white space between clusters, which may better reflect the global relations between clusters.

@@

We introduce \opentsne, a comprehensive Python library that implements the
t-SNE algorithm and its numerous recentl extensions. \opentsne\ is
designed to accomodate all types of users and includes both a simple API that
allows users to quickly obtain informative t-SNE visualizations, as well as a
more flexible API that allows seasoned practitioners to modify all the
different aspects of the t-SNE algorithm.

The t-SNE algorithm is comprised of three main steps, mirrored in the structure
and flow of the \opentsne\ library. These are: the initial positions
of points in the embedding, an affinity model describing the similarities
between data points in the embedding, and the optimization phase which
optimizes the embedding layout. Using the advanced API, each of these key
elements can be modified or tuned according to produce the best possible
visualization.

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{opentsne-diagram}
  \caption{\label{fig:diagram}
  Bla
  }
\end{figure*}

\opentsne\ is compatible with the Python data science ecosystem (e.g. \pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar and intuitive API to end-users. Its modular design encourages extensibility and experimentation with various parameter settings and changes to the analysis pipeline. \opentsne\ aims to make the latest theoretical advances in t-SNE accessible to the wider data science community. These features are easily accessible through an intuitive API, which closely follows the style \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit in the machine learning community.
The following snippet exemplifies this interface and demonstrates how \opentsne\ can be used to quickly generate several visualizations, each highlighting different aspects of the underlying data structure.

\subsection{Example}
\subsubsection{Installation}

\opentsne\ is designed to be as user-friendly as possible and can be installed through either \pkg{PyPI} using \pkg{pip}
\begin{CodeChunk}
\begin{CodeInput}
pip install opentsne
\end{CodeInput}
\end{CodeChunk}
or using \pkg{conda} from the \pkg{conda-forge} channel
\begin{CodeChunk}
\begin{CodeInput}
conda install -c conda-forge opentsne
\end{CodeInput}
\end{CodeChunk}
\opentsne\ provides precompiled binaries for all major platforms (Windows, MacOS, Linux) for all currently supported Python versions (3.6---3.10). This is especially convenient for Windows users, as this obviates the need for a C/C++ compiler, which do not come bundled with the operating system.

To verify that that \opentsne\ was installed correctly, try to import the module:
\begin{CodeChunk}
\begin{CodeInput}
import openTSNE
print(openTSNE.__version__)
\end{CodeInput}
\end{CodeChunk}

\subsubsection{A simple workflow}

\begin{CodeChunk}
\begin{CodeInput}
from openTSNE import TSNE

@@

\section{Discussion and Examples} \label{sec:summary}

@@

\subsection{Embedding Massive Data Sets}

When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between clusters become increasingly difficult to interpret. We constructed Fig.~\ref{fig:cao}.a from the data containing expression profiles of over two million single cells captured at different time points in mouse development. The embedding indicates numerous clusters with transitions between time points, as marked by the color-coding, that are difficult to interpret. Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently, more informative visualizations~\citep{kobak2019art}. For instance, \ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the data's overall structure. Further doubling the exaggeration in \ref{fig:cao}.c allows us to observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters. The visualization also reveals several tiny clusters, possibly corresponding to rare cell types.
\begin{CodeChunk}
\begin{CodeInput}
# To improve the global structure of the final layout, we initialize the
# embedding by creating a t-SNE plot on a sample of points

@@

# We can next use the `transform` functionality to determine starting positions
# for the remaining data points. Since we don't actually want to optimize the
# remaining points separately from the sample points, we don't need to call the
# `transform` function. Instead, we'll call the `prepare_partial` function,
# which `transform` calls anyways under the hood.
init_rest = init_sample.prepare_partial(x_rest)

# Lastly, we merge `init_sample` and `init_rest` and restore the original order
init_full = np.vstack((init_sample, init_rest))[reverse]
\end{CodeInput}
\end{CodeChunk}
We generate the three different embeddings using the following code snippet (which is shortened for clarity).
\begin{CodeChunk}
\begin{CodeInput}
# We separately precompute the affinities as this can take quite a long time.
# This way, we can reuse the precomputed affinities in various different
# embeddings, without having to repeat the same, costly computation every time.
affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)

@@

% General problem statement
One of the t-SNE's common criticisms is limited scalability to large data sets
containing, for instance, millions of data points~\citep{becht2019dimensionality}.
The culprit for slow response time stems from
particular optimization procedures and their specific implementation in popular
open-source libraries. Until quite recently, most popular implementations of
t-SNE were based on either a direct implementation of the t-SNE algorithm with
asymptotic time complexity $\mathcal{O}(N^2)$, or the Barnes-Hut approximation
scheme developed by
\citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}
 (N \log N)$, where $N$ is the number of data items ({\em e.g.} cells).

% Problem with Python
In Python, the most widely-used implementation of t-SNE came from \pkg{scikit-learn},
which exhibits long runtimes when compared to its C++ counterpart --
\pkg{MulticoreTSNE}. The multi-threaded \pkg{MulticoreTSNE}
implementation~\citep{Ulyanov2016} can construct t-SNE embeddings of millions of
data points in a matter of hours (Fig.~\ref{fig:benchmarks_py}).
However, recently, \citet{linderman2019fast} developed a
new approximation scheme -- FIt-SNE -- which futher reduces the asymptotic time
complexity to $\mathcal{O}(N)$. This approximation scheme is included in \pkg{FIt-SNE} and \opentsne, enabling the embedding of large data sets in a matter
of minutes.

% We benchmark Python
Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python t-SNE implementations,
including those from \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1),
\pkg{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.6.1). Benchmarks were performed on
an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original
t-SNE parameters, as some implementations do not allow for their modification.

To ensure a realistic benchmarking scenario, we utilize the 10X
Genomics 1.3 million mouse brain data set, containing 1.3 million single-cell 
gene-expression profiles. We preprocess the data using the standard single-cell
analysis pipeline and extract the top 50 principal components. To generate
benchmark data sets of different sizes, we subsample the data set five times,
resulting in five different data sets for each sample size. In total, we run
each implementation on 30 different data sets.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{benchmarks_python-final}
  \caption{\label{fig:benchmarks_py}
  We benchmark \opentsne\ (v0.6.1) against three popular open-source
  implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
  (v1.0.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
  on a server-grade Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
  that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
  same interpolation-based approximation scheme, while \pkg{scikit-learn} and
  \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
}
\end{figure*}

@@

% We benchmark other programming languages
Using the same benchmarking setup as before, we ...
The results are shown in Fig.~\ref{fig:benchmarks_lang}. 
Using 8 cores, Rtsne runs out of stack memory at 500k

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{benchmarks_langs-final}
  %\includegraphics[width=\textwidth]{benchmarks-combined}
  \caption{\label{fig:benchmarks_lang}
  We choose the fastest available, open-source implementations of the t-SNE algorithm from the Python, R, and Julia programming languages. We choose \opentsne\ for Python, \pkg{Rtsne} for R, and \pkg{TSne.jl} for Julia. The benchmarks reflect the time complexity of the implemented algorithms. For instance, \pkg{TSne.jl} implements a naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all but the smallest data sets. \pkg{Rtsne} scales similarily to \pkg{MulticoreTSNE} in Python, as it implements the $\mathcal{O}(N \log N)$ Barnes-Hut approximation scheme. Interestingly, the multi-threaded implementation version of \pkg{Rtsne} crashes on benchmark data sets containing over 500,000 data points due to lack of memory. This would seriously hamper its utility on consumer-grade platforms, as our benchmark platform contained 128 GB of memory. \opentsne\ is the only library implementing the more recent FIt-SNE approximation scheme, which makes it suitable for even the largest data sets.
}
\end{figure*}

@@

We designed \opentsne\ to be accessible to a broader audience. We provide
precompiled binaries for all major Python versions on all major platforms,
making the installation process as seamless as possible. One can install
\opentsne\ through the Python Package Index (\pkg{PyPI}) or \pkg{conda} from the
\pkg{conda-forge} channel, the two most widely adopted Python package managers.
\opentsne's high-level interface is inspired by \pkg{scikit-learn}, which is well
established in the Python data science ecosystem. \opentsne\ implements
multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation
schemes, enabling it to be applied to data sets containing millions of data
points. While the Python virtual machine inevitably introduces some performance
overhead, the runtime is comparable to its C++ counterpart -- \pkg{FIt-SNE}.
Finally, \opentsne\ is extensible. Its modular design enables researchers to
quickly experiment with different parameter settings and easily incorporate
custom components into the software. We provide an extended list of features and
comparison to other popular t-SNE implementations in Table~\ref{tab:features}.
